\iffalse
One-page printable interview handout for
"Communication-Efficient Learning of Deep Networks from Decentralized Data"
arXiv:1602.05629  (McMahan et al., AISTATS 2017 -- FedAvg)
Conforms to write-latex rules: no fontspec, \section* unnumbered, pdflatex-clean,
comments only inside \iffalse...\fi. NO \citep -- this file loads no natbib; all
references are inline plaintext.
\fi

\documentclass[10pt]{article}

\usepackage[letterpaper, margin=0.6in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{titlesec}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  citecolor=blue!50!black,
  urlcolor=blue!50!black
}

\titleformat{\section}{\large\bfseries}{}{0pt}{}
\titlespacing*{\section}{0pt}{6pt}{2pt}

\setlist[itemize]{leftmargin=*, topsep=1pt, itemsep=1pt}

\pagenumbering{gobble}

\begin{document}

\begin{center}
  {\Large\textbf{McMahan et al.\ 2017 --- Communication-Efficient Learning from Decentralized Data (FedAvg)}}\\[2pt]
  {\small Federated Learning \& the FederatedAveraging algorithm \quad\textbar\quad arXiv:1602.05629 \quad\textbar\quad \href{https://github.com/pleyva2004/communication-efficient-learning-of-deep-networks}{my study repo}}
\end{center}

\section*{One-Liner}
Train a shared model on data that never leaves the device: each client runs several epochs of local SGD, the server averages the resulting \emph{models}; this cuts communication rounds $10$--$100\times$ versus synchronized SGD.

\section*{What's Novel}
\begin{itemize}
  \item The \emph{problem} and its framing --- ``federated optimization'' defined by four properties (non-IID, unbalanced, massively distributed, limited communication) --- not the averaging algorithm, which predates the paper (McDonald et al.\ 2010; Povey et al.\ 2015).
  \item One dominant knob: spend ``essentially free'' on-device computation ($E$ local epochs, minibatch $B$) to buy the scarce resource, communication rounds. Speedups come mostly from more local work, not more clients.
  \item An evaluation methodology for the regime that violates every assumption of convex distributed optimization ($K\!\ll\!n_k$, IID, equal $n_k$).
\end{itemize}

\section*{Math in 60 Seconds}
The objective is a finite sum, decomposable exactly over clients (no IID needed):
$$ f(w)=\tfrac{1}{n}\textstyle\sum_i f_i(w)=\sum_{k}\tfrac{n_k}{n}F_k(w). $$
Differentiating gives $\sum_k \tfrac{n_k}{n}g_k=\nabla f(w_t)$, so \textbf{FedSGD} ($B{=}\infty,E{=}1,C{=}1$) is exactly full-batch GD. Because $\sum_k n_k/n=1$, averaging one local step equals one global step --- \textbf{FedAvg} just iterates the local step first. For $E>1$ the per-round gap from centralized GD is, to leading order, $\Delta=\eta^2\,\mathrm{Cov}_k(H_k,g_k)$: a pure client-\emph{heterogeneity} term (zero when IID). Non-convex averaging only works from a \emph{shared} init (Figure 1: independent inits cross a loss barrier); FedAvg re-broadcasts $w_t$ each round to guarantee it.

\section*{What I'd Push Back On}
\begin{itemize}
  \item No theory at all --- no convergence result, no rate; the $E>1$ regime is uncharacterized and can ``plateau or diverge.'' The justification for averaging is one 1-D loss-interpolation plot.
  \item Evaluation flatters the method: per-$x$-point learning-rate grid search ($>2000$ models), monotonized curves, interpolated round counts, no error bars; the FedSGD baseline (one full-batch step/round) is the weakest possible comparator.
  \item ``Computation is free'' is asserted, never measured (up to $1200$ local updates/round on a phone); the headline $95\times$ is the \emph{unbalanced} Shakespeare case the authors admit is easier, and the erratum's $n_k/m_t$ average is only exactly unbiased when clients are balanced.
\end{itemize}

\section*{My One Proposed Extension}
Cancel the heterogeneity drift directly. Since the $E>1$ ``free lunch'' is exactly the covariance $\eta^2\mathrm{Cov}_k(H_k,g_k)$, add SCAFFOLD-style control variates so each local step uses $g_{\mathrm{local}}-c_k+c$, re-centering clients onto the global direction at no extra backward pass (SCAFFOLD ``Option II''). My prototype recovers $1.62\times$ fewer rounds on the pathological non-IID partition --- operationalizing the paper's own weakest spot, and orthogonal to the aggregation-bias fix, the $E$-decay schedule, and permutation-aligned averaging.

\emph{Full proposal set: \texttt{05-improvements.tex} in the study repo.}

\section*{My One Question}
\emph{For $E>1$ under genuine non-IID data, what fixed point (if any) does FedAvg converge to, and is there a principled rule for choosing $(E,B)$ --- given that Table 2 shows more local work sometimes makes things \emph{worse}?}

\end{document}
